---
title: "Exercise 6"
author: "Gerko Vink"
date: "Fundamental Techniques in Data Science with R"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
  
<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>
  
---
  
```{r echo=FALSE}
printAnswers <- FALSE
```

# Assumptions

---

## Inspect `fit1`

---

8. **Inspect the assumptions for the first model. What do you think?**

HINT: use `plot()` and use the plots you've created in exercises 5-7.

```{r}
plot(fit1)
```

- The data follows a linear trend, although the loess curve shows some deviations from linearity. All in all, there are only 11 points, so this slight deviations is not something that would worry me. 
- Taking into account that there are only a few observations, I would argue that the residuals seem normally distributed from the `Normal Q-Q` plot. 
- The residual variance seems constant over the level of the fitted values (i.e. homoscedastic residual variance) as seen in `Residuals vs. Fitted` plot and the `Scale-Location` plot. Again, the dip in the `Scale-Location` plot can easily be explained by the small sample size and the deviation should be taken with a grain of salt. 
- No special remarks with respect to leverage and cook's distance, although case #3 would need to be at least investigated. 

---

## Inspect `fit2`

---

9. **Inspect the assumptions for the model `fit2`. What do you think?**

```{r echo=printAnswers, eval = printAnswers}
#- The data does not follow a linear trend, the deviation would definitely worry me. 
#- The residuals seem non-normally distributed, especially in the tails from the `Normal Q-Q` plot. 
#- I could not still argue that the residual variance seems more-or-less constant over the level of the fitted values. The residual variance is heteroscedastic.  
#- Case 8 has quite some leverage and a large residual. It's cook's distance is greater than `.5`. 

plot(fit2)
```

---

## Inspect `fit3`

---

10. **Inspect the assumptions for the model `fit3`. What do you think?**

```{r echo=printAnswers, eval = printAnswers}
#- The data follows a perfect linear trend, except for case #3 
#- The residuals seem normally distributed, except for case #3
#- I could not still argue that the residual variance seems more-or-less constant over the level of the fitted values. The residual variance is heteroscedastic. However, if case #3 were omitted, there are no residuals: every point falls perfectly on the regression line. 
#- Case 3 has quite some leverage, but not as large as other cases. Case #3 has the largest residual. It's cook's distance is greater than `1`. 

plot(fit3)
```

---

## Inspect `fit4`

---

11. **Inspect the assumptions for the model `fit4`. What do you think?**

```{r echo=printAnswers, eval = printAnswers}
#- The data follows no trend. You'd be an idiot to perform linear regression on this set. 
#- The residuals seem normally distributed
#- I could not still argue that the residual variance seems more-or-less constant over the level of the fitted values. The residual variance does not exist for fitted values other than 7!
#- Case 8 has a leverage of 1; hence it is omitted from most of the plots. The plot over the remaining points is redundant.  

plot(fit4)
```

