
---
title: Problem Set FTID_1
output: 
  html_document: 
    keep_md: yes
    toc: yes
---

```{r setup, include=FALSE, echo=FALSE}
# Load libraries and source extra code
library(magrittr)
library(dplyr)
library(ggplot2)
library(RTutor)


# Options for rendering data frames
# If you knit to a Word docx file, try
# 
# data.frame.theme="word" 
# 
# (needs RStudio > 1.2.1)
# 
# You can also set the options like
# table.max.cols as chunk options
# Makes sense if there are too many, too wide
# columns in some chunks

RTutor::set.knit.print.opts(data.frame.theme="code", table.max.rows=25, table.max.cols=NULL, round.digits=5, signif.digits=8)


# continue knitting even if there is an error
knitr::opts_chunk$set(error = TRUE) 
```

# R: Assigning elements to objects

Author:  Gerko Vink


In this exercise set we will get acquainted with `R`. If you have no experience with `R`, you will learn the most from following this document by looking at the solutions. If you have some experience with `R` already, I suggest you try solving the exercises without looking at the solutions. You can then refer to the solutions at any time, if needed. 

If you have any questions or if you feel that some code should be elaborated, feel free to ask during the workgroup or drop me a line at [G.Vink@uu.nl](mailto::G.Vink@uu.nl).

Have fun at finishing the exercises.

## Exercise content

1. Basic operations in R

  1.1 Assigning elements to objects

  1.2 Simple calculations with elements and objects

2. Objects with more than one element

  2.1 Vectors

  2.2 Matrices

  2.3 Data Frames

  2.4 Arrays

  2.5 Lists

## Exercise 1 -- Assigning elements to objects

---

We start with the very basics and will move towards more advanced operations in `R`. First we will get acquainted with the language (which is called `S` - to keep it simple) and the environment we work in. 

---

## Exercises 1.1 -- Simple assigning

`R` is an object oriented language. This means that elements, such as numbers or characters, are assigned to objects. The below code block assigns the value `1` to object `a`. 

```{r "1",optional=TRUE, eval=FALSE}
a <- 1
```

The `#` tells `R` that everything that follows in that specific line is not to be considered as code. In other words, you can use `#` to comment in your own `R`-scripts. I used `#` here to elaborate that the following line is the code from exercise 1. 

The line `a <- 1` assigns the value `1` to object `a`. When you run your code, it will be saved. The value 1 and the letter `a` are chosen to illustrate assigning in `R`. You might as well assign `123` to `banana` if you like. Really, anything goes. 
