---
title: "Exercise 4"
author: "Gerko Vink"
date: "Fundamental Techniques in Data Science with R"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
    font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>



```{r echo=FALSE}
printAnswers <- TRUE
```


---

# Exercises

---

The following packages are required for this practical:
```{r, message=FALSE}
library(dplyr)
library(magrittr)
library(mice)
```
and if you'd like the same results as I have obtained, you can fix the random seed
```{r}
set.seed(123)
```

---

## Exercise 1-5

---

1. **Use a pipe to do the following:**

- draw 1000 values from a normal distribution with `mean = 5` and `sd = 1` - $N(5, 1)$, 
  - HINT: use `rnorm()`
- create a matrix where the first 500 values are the first column and the second 500 values are the second column
- make a scatterplot of these two columns
```{r}
rnorm(1000, 5) %>%
  matrix(ncol = 2) %>%
  plot()
```

---

2. **Use a pipe to calculate the correlation matrix on the `anscombe` data set**

```{r}
anscombe %>%
  cor()
```

---

3. **Now use a pipe to calculate the correlation for the pair (`x4`, `y4`) on the `anscombe` data set**

Using the standard `%>%` pipe:
```{r}
anscombe %>%
  subset(select = c(x4, y4)) %>%
  cor()
```
Alternatively, we can use the `%$%` pipe from package `magrittr` to make this process much more efficient.
```{r}
anscombe %$%
  cor(x4, y4)
```

---

4. **Use a pipe to calculate the correlation between `hgt` and `wgt` in the `boys` data set from package `mice`.**

Because `boys` has missings values for almost all variables, we must first select `wgt` and `hgt` and then omit the rows that have missing values, before we can calculate the correlation. Using the standard `%>%` pipe, this would look like:
```{r}
boys %>%
  subset(select = c("wgt", "hgt")) %>%
  cor(use = "pairwise.complete.obs")
```
which is equivalent to 
```{r}
boys %>%
  subset(select = c("wgt", "hgt")) %>%
  na.omit() %>%
  cor()
```

Alternatively, we can use the `%$%` pipe:
```{r}
boys %$% 
  cor(hgt, wgt, use = "pairwise.complete.obs")
```
The exposition `%$%` pipe *exposes* the listed dimensions of the `boys` dataset, such that we can refer to them directly. 

---

5. **In the `boys` data set, `hgt` is recorded in centimeters. Use a pipe to transform `hgt` in the `boys` dataset to height in meters and verify the transformation**

Using the standard `%>%` and the `%$%` pipes:
```{r}
boys %>%
  transform(hgt = hgt / 100) %$%
  mean(hgt, na.rm = TRUE)
```

---

## Exercise 6-8

---

6. **Use pipes to plot the pair (`hgt`, `wgt`) two times: once for `hgt` in meters and once for `hgt` in centimeters. Make the points in the 'centimeter' plot `red` and in the 'meter' plot `blue`. **
```{r}
boys %>%
  subset(select = c(hgt, wgt)) %>%
  plot(col = "red", main = "Height in centimeters") 
boys %>%
  subset(select = c(hgt, wgt)) %>%
  transform(hgt = hgt / 100) %>%
  plot(col = "blue", main = "Height in meters")
```

---

```{r echo=FALSE}
set.seed(32083)
x <- rnorm(100, mean=3, sd=7)
y <- c(rnorm(50, mean = 2, sd = 3), rnorm(50, mean = 9, sd = 2))
save(x, y, file = "Exercise4_data.RData")
```


In the following experiment we investigate least-squares estimation of the mean. The data for this experiment can be found in the workspace [`Exercise4_data.RData`](Exercise4_data.RData). Either download this workspace and manually `load()` it into `R`, or run the below connection:

```{r eval=FALSE}
con <- url("https://www.gerkovink.com/fundamentals/Wk2/Exercise/Exercise4_data.RData")
load(con)
```

The workspace contains a single vector, named `x` and a vector named `y`

---

7. **Obtain the sample mean of the values in `x`.**
```{r}
mean(x)
```

---

The values in `x` have been drawn from a population with mean $\mu = 3$ and standard deviation $\sigma = 7$. 

---

8. **Calculate the sample mean's sum of squared deviations from $\mu$. The sum of squared deviations from mu is defined as: **
$$ \text{ssd} = \sum_{i=1}^{100} (x_i - \mu)^2.$$
There is a slow way
```{r}
mu = 3
deviations <- x - mu
ssd <- sum(deviations^2)
```

And a fast way with function apply, where a function is applied over the margin (1 = rows, 2 = columns) of some data. 
```{r}
ssd2 <- apply(X = outer(x, mu, FUN = "-")^2, 
              MARGIN = 2, 
              FUN = sum)
```
Here `X = outer(x, mu, FUN = "-")^2` provides the data to apply. Remember that we have defined `x` and `mu` in the global environment. With the `outer()` function, we can quickly determine the outer product of some vectors. In this case we simply use subtraction `-` as our function to obtain for the column margin (`MARGIN = 2`) - which in this case is just the vector of squared deviations - and sum (`FUN = sum`) all these values into a single sum. The result is the sum of squared deviations. 

Both solutions are identical
```{r}
identical(ssd, ssd2)
```

To see them, we can simply congatenate the two values:
```{r}
c(ssd, ssd2)
```

---

# Functions in `R`

The apply class of functions is very flexible and lightning fast, when compared to manual operations that could easily be defined in terms of functions. The only caveat is that you need a function to apply. Many such functions are already available in `R`, such as `mean()`, `mode()`, `sum()`, `cor()`, and so on. 

However, if you need to perform more than a simple calculation, it is often necessary to create your own function. In `R` functions take the following form

```{r}
myfunction <- function(arguments){
  hereyourfunctioncode
}
```

For example, 
```{r}
mean.sd <- function(argument1, argument2){
  mean1 <- mean(argument1) 
  mean2 <- mean(argument2)
  sd1 <- sd(argument1)
  sd2 <- sd(argument2)
  result <- data.frame(mean = c(mean1, mean2),
                       sd = c(sd1, sd2), 
                       row.names = c("first", "second"))
  return(result)
}
```
The above function calculates the means and standard deviations for two sources of input, then combines these statistics in a simple data frame and returns the data frame. The sources of input are defined in the function arguments `argument1` and `argument2`. The reason why we have to specify these arguments is simple:

\[\text{EVERYTHING THAT HAPPENS IN A FUNCTION COMES FROM THE}\] 
\[\text{FUNCTION AND STAYS IN THE FUNCTION!}\]

This is because a function opens a seperate environment that only exists for as long as the function operates. This means:

1. To get information from the global environment to the function's environment, we need arguments.
2. To properly return information to the global environment, we should use `return()`. In general, using `return()` makes it explicit what your function's return is. For complicated functions this is proper coding procedure, but for simple functions it is not strictly necessary. 

To put this example function to the test:
```{r}
mean.sd(argument1 = 1:10,
        argument2 = 3:8)
```

or, simply:
```{r}
mean.sd(1:10, 3:8)
```


---

## Exercises 9-10
9. **Now create a function that automates the calculation of the sum of squares for any given $\mu$. Call the function `lsfun` because we are going to identify the least squares estimate and coding is fun!**

We can use the 
```{r eval=printAnswers}
lsfun <- function(meanestimate) apply(outer(x, meanestimate, "-")^2, 2, sum)
```

or, 100% equivalently, but easier to spot as a function:
```{r}
lsfun2 <- function(meanestimate){
  apply(outer(x, meanestimate, "-")^2, 
        MARGIN =  2, 
        FUN = sum)
}
```

or, with a pipe
```{r}
lsfun3 <- function(meanestimate){
  outer(x, meanestimate, "-")^2 %>%
    apply(MARGIN = 2, FUN = sum)
}
```

---

10. **Plot the curve of your least square function such that you can identify the minimum of the curve (i.e. the location for $x$ where the sum of the squared deviations are the lowest).**
```{r}
curve(lsfun, from = 1, to = 8, ylab = "Sum of the squared deviations")
```

We can see that the minimum lies somewhere above, but near the value 4. We already know from exercise 7 that the mean is equal to `r mean(x)`, and that is the `x` coordinate for the abscissa. To verify this, we can change the range of the plotted curve to zoom in on `r mean(x)`:

```{r}
curve(lsfun, from = 4.16, to = 4.17)
```


---


11. **Repeat the experiment from 10 on object `y` from `Exercise4_data.RData`.**

- Plot a histogram of `y` and report on the shape of the data
  - What do you think the mean is, based on the plot?
  - Calculate the mean (expected value) 
  - Calculate the median (the center of the distribution) 
  - Calculate the mode (the most observed value). Calculate the mode on rounded numbers (HINT: use `round()`).
  - what do the mean, mode and median tell you about the shape of the data?
- Plot the least squares curve from values 2 to 10, conform the previous exercise
- Plot a curve that zooms in on the minimum (i.e. the least squared deviations)
- Report the `meanestimate` that would minimize the least squares function

```{r eval = FALSE, echo = FALSE}
hist(y)
mean(y)
median(y)
mode <- round(y) %>% table 
# mode is equal to 10
lsfun <- function(meanestimate) apply(outer(y, meanestimate, "-")^2, 2, sum)
curve(lsfun, from = 2, to = 10)
curve(lsfun, from = 2, to = 10)
```

---

End of practical. 

 
---

# Useful References

- [`magrittr`](https://magrittr.tidyverse.org)
- [`R` for Data Science](http://r4ds.had.co.nz) - [Chapter 18 on pipes](http://r4ds.had.co.nz/pipes.html)