---
title: "Exercise 4 - answers"
author: "Gerko Vink"
date: "Fundamental Techniques in Data Science with R"
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---

---


```{r echo = FALSE}
library(magrittr) #pipes
con <- url("https://www.gerkovink.com/fundamentals/Wk2/Exercise/Exercise4_data.RData")
load(con)
```

---


11. **Repeat the experiment from 10 on object `y` from `Exercise4_data.RData`.**

- Plot a histogram of `y` and report on the shape of the data
```{r}
hist(y)
```

  - What do you think the mean is, based on the plot?
  
**Hard to say, the data seems to be bi-modal. Somewhere around 6? as there seems to be more mass around 10 than around 1.**

  - Calculate the mean (expected value) 
```{r}
mean(y)
```
  - Calculate the median (the center of the distribution) 
```{r}
median(y)
```
  - Calculate the mode (the most observed value). Calculate the mode on rounded numbers (HINT: use `round()`).
```{r}
round(y) %>% table
```
**The mode seems to be around 10**
  
  - what do the mean, mode and median tell you about the shape of the data?
  
**The mean, median and mode are all different. This means data de data are non-normal. The histogram clearly shows this. If we plot the density, this becomes even more apparent:**
```{r}
y %>% density %>% plot
```

- Plot the least squares curve from values 2 to 10, conform the previous exercise
```{r}
lsfun <- function(meanestimate) apply(outer(y, meanestimate, "-")^2, 2, sum)
curve(lsfun, from = 2, to = 10)
```

- Plot a curve that zooms in on the minimum (i.e. the least squared deviations)
```{r}
curve(lsfun, from = 5.5, to = 6)

```

- Report the `meanestimate` that would minimize the least squares function

**Naturally, this is the mean of `y`, which equals `r mean(y)`**

---

End of practical. 

 
---

# Useful References

- [`magrittr`](https://magrittr.tidyverse.org)
- [`R` for Data Science](http://r4ds.had.co.nz) - [Chapter 18 on pipes](http://r4ds.had.co.nz/pipes.html)